openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-07 paper/policar2020-opentsne.tex
Minor corrections

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new samples into existing
	embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multiscale similarity kernels~\cite{kobak2019art} which preserves small,
	well-defined clusters and uncovers global relationships between clusters,
	and
\item improved defaults for learning rate and the number of
	iterations~\cite{belkina2019automated} which produce embeddings at a lower
	computational cost.
\end{enumerate}

The library is fully compatible with the Python data-science ecosystem (e.g.,
\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
and intuitive user interface to new users. Its modular design encourages
extensibility and experimentation with various settings and changes in the
analysis pipeline.

@@

Accessibility is one of the core design principles of \opentsne. Our aim is to
make the latest theoretical advances in t-SNE as accessible as possible to as
many users as possible. As such, our API is based on the familiar
\textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a
smooth transition from existing Python libraries. For example,
Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
\opentsne, each based on recent advances in the t-SNE algorithm. The code used
to generate these embeddings is shown below.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence to suggest that
any data set can accurately be described using only two dimensions and that any
projection onto a two-dimensional plane will inevitably lead to loss of
information. Therefore, it is beneficial to examine multiple embeddings, each
of which provides a different perspective into the topology and other
characteristics of the data.
\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data
	set containing 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
or instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which
uncovers numerous clusters. Warm colors correspond to excitatory neurons, while
cool colors represent inhibitory neurons. While different clusters of
excitatory and inhibitory neurons appear close to one another, all clusters
appear somewhat equidistant from their neighboring clusters, and the overall
relations between them are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white-space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups.

When faced with a data set containing millions of data points, standard t-SNE
embeddings often become unwieldy, with blurred cluster boundaries, large
clusters absorbing smaller ones, and relationships between clusters becoming
increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
profiles of over two million single cells captured at different time points in
mouse development. The embedding indicates numerous clusters with some
transition between time points as indicated by the color-coding, but this is
largely difficult to interpret. Kobak \& Berens observed that increasing
attractive forces between similar data points controlled via the
\textit{exaggeration} parameter leads to more compact clusters, and
subsequently more informative visualizations~\cite{kobak2019art}. For instance,
\ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the
overall structure of the data. Doubling the exaggeration again in
\ref{fig:cao}.c, we can now easily observe that the data is comprised of two
main groups of cells and eight somewhat smaller clusters. The visualization
also reveals a small number of very small clusters, possibly corresponding to
rare cell types.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}The data set from Cao \textit{et
	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
	single cells. The data were collected from mice embryos at different
	developmental time-points at daily intervals after 9.5 to 13.5 days.
	The exaggeration parameter can be used to highlight the global
	organization of the clusters and enforce cluster compactness.
	\textbf{(a)} shows the data is comprised of two main components -- the
	neural tube and mesenchymal cells -- as well as several other smaller
	clusters. Additionally, the colors indicate developmental progression
	with red indicating least-developed cells and blue indicating most
	developed cells. The overall developmental trajectory is most apparent
	with higher levels of exaggeration, showing red cells slowly
	transitioning into blue cells. Progressively easing the exaggeration
	factor uncovers finer clusters within the larger groups as shown in
	\textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)},
	where we show the standard t-SNE with no exaggeration. 32,011 putative
	doublets are excluded from the visualizations.}
\end{figure*}

Exaggeration can be used to highlight transitions between cell-states in
developmental studies. Standard t-SNE often produces embeddings with
clearly-defined, discrete clusters. We have shown that we can adjust the level
of granularity and resolution of the clusters with several parameters. However,
discrete clusters are often undesired in developmental studies where we assume
cells follow a continuous path between cell states. To this end, other
embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2}
are often used, as they tend to better capture the continuity between cell
states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
in Fig.~\ref{fig:cao}.a, the developmental path between different time points
is difficult to observe due to a large number of sprawled out clusters in the
embedding. On the other hand, the developmental paths are easy to follow
between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
factor to $4$.

@@

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases where cell identities are not known in
advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
embedding using annotated data from Hochgerner \textit{et
al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
cells from the mouse brain. The authors provide annotations for each cell, and
we can verify their accuracy by visualizing well-established gene markers'
major cell types. We then embed cells from Harris \textit{et
al.}~\cite{harris2018classes}, where annotations are provided only for neuronal
cells. We quickly identify other non-neuronal cell types, including
oligodendrocytes and astrocytes. We use marker genes to validate that new cells
were correctly mapped into the reference landscape.

@@

These two examples demonstrate how \opentsne\ can be used to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing already
existing, publicly-available, reference atlases. This approach is extremely
general and is not limited to single-cell gene-expression data, but can be
applied to any tabular data sets from any domain of science.

@@

Versatility is another of  \opentsne s core design principles. Kobak \& Berens
recently introduced several recommendations and tricks to obtain better and
more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
of multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with larger data sets. \opentsne\ provides an
extremely flexible API, where these improvements can be incorporated in just a
few lines of code. Furthermore, \opentsne\ supports custom affinity models,
e.g. multi-scale Gaussian mixtures or uniform kernels, allowing for quick
experimentation. This also enables users to construct t-SNE embeddings on
non-tabular data, e.g. relational data, as the only affinity-model requirement
is some notion of similarity between data points. Finally, our comprehensive
callback system can be utilized to monitor and adapt different stages of the
optimization phase and has been used to construct visually appealing animations
of the t-SNE optimization process.

@@

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
\opentsne\ exhibit better runtime performance than their Barnes-Hut
counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
even though \textsf{FIt-SNE} and \opentsne\ both implement the same
approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
written in Python, which inevitably incurs some runtime overhead compared to
its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} demonstrates the
runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
than its Barnes-Hut competitors. Additionally, modern computer processors
generally contain multiple cores, allowing us to make use of multi-threading.
This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart. We speculate that this is could be because \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which are aggressively optimized on
Intel(R) Xeon(R) processors.

@@

This section is meant to familiarize the reader with the t-SNE algorithm and
its recent extensions. This brief review is by no means comprehensive, but is
self-contained and introduces the mathematics upon which the \opentsne\ library
is based.

>>>>

We introduce \opentsne, a comprehensive Python library that implements t-SNE
and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation 
	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
	of millions of data points,
\item the addition of new samples into existing
	embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more
	globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
	of data at different levels of resolution,
\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
	well-defined clusters and uncover global relationships between clusters,
	and
\item improved defaults for learning rate and number of
	iterations~\cite{belkina2019automated} which produce embeddings at a lower
	computational cost.
\end{enumerate}

The library is fully compatible with the Python data-science ecosystem (e.g.,
\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
and intuitive API to new users. Its modular design encourages
extensibility and experimentation with various settings and changes in the
analysis pipeline.

@@

Accessibility is one of the core design principles of \opentsne. Our aim is to
make the latest theoretical advances in t-SNE as accessible as possible to as
many users as possible. As such, our API is based on the familiar
\textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a
smooth transition from existing Python libraries. For example,
Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
\opentsne, each based on recent theoretical advances in the t-SNE algorithm. The code used
to generate these embeddings is shown below.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional
data lies on a lower-dimensional manifold, which can accurately be captured by
a smaller number of dimensions. However, there is no evidence to suggest that
any data set can accurately be described using only two dimensions and that any
embedding onto a two-dimensional plane will inevitably lead to loss of
information. Therefore, it is beneficial to examine multiple embeddings, each
of which provides a different perspective into the topology and other
characteristics of the data.
\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use openTSNE to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
Warm colors correspond to excitatory neurons, while
cool colors represent inhibitory neurons. While different clusters of
excitatory and inhibitory neurons appear close to one another, all clusters
appear somewhat equidistant from their neighbors, and the overall
relations between groups are not obvious. On the other hand,
Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
resulting in a more globally consistent layout where relations between clusters
become more apparent. In this embedding, it is evident from the increased
white-space between groups that there is one large class of excitatory neurons,
and two related classes of inhibitory neurons. Unfortunately, focusing on
preserving large neighborhoods often results in smaller clusters being absorbed
into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
similarity kernels which aim to preserve both the global organization of
clusters as well as prevent smaller clusters from being absorbed into larger
ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
reveals that many of the clusters are composed of other subgroups and may be
clustered further into cell sub-types.

When working with data containing millions of data points, standard t-SNE
embeddings often become unwieldy -- cluster boundaries become blurred, large
clusters absorb smaller ones, and relationships between clusters become
increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
profiles of over two million single cells captured at different time points in
mouse development. The embedding indicates numerous clusters with some
transition between time points as indicated by the color-coding, but this is
largely difficult to interpret. Kobak \& Berens observed that increasing
attractive forces between similar data points controlled via the
\textit{exaggeration} parameter leads to more compact clusters, and
subsequently more informative visualizations~\cite{kobak2019art}. For instance,
\ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the
overall structure of the data. Doubling the exaggeration again in
\ref{fig:cao}.c, we can now easily observe that the data is comprised of two
main groups of cells and eight somewhat smaller clusters. The visualization
also reveals a number of tiny clusters, possibly corresponding to
rare cell types.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}Increasing the exaggeration parameter produces more
  	compact clusters, which highlight the global organization of the clusters
  	and emphasizes continuities between cell states. The data set from Cao \textit{et
	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
	single cells. The data were collected from mice embryos at different
	developmental time-points at daily intervals after 9.5 to 13.5 days.
	\textbf{(a)} shows the data is comprised of two main components -- the
	neural tube and mesenchymal cells -- as well as several other smaller
	clusters. Additionally, the colors indicate developmental progression
	with red indicating least-developed cells and blue indicating most
	developed cells. The overall developmental trajectory is most apparent
	with higher levels of exaggeration, showing red cells slowly
	transitioning into blue cells. Progressively easing the exaggeration
	factor uncovers finer clusters within the larger groups as shown in
	\textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)},
	where we show the standard t-SNE with no exaggeration. 32,011 putative
	doublets are excluded from the visualizations.}
\end{figure*}

Exaggeration can highlight transitions between cell-states in
developmental studies. Standard t-SNE often produces embeddings with
clearly-defined, discrete clusters. We have shown that we can adjust the level
of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
discrete clusters are often undesired in developmental studies where we assume
cells follow a continuous path between cell states. To this end, other
embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2}
are often used, as they tend to better capture the continuity between cell
states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
in Fig.~\ref{fig:cao}.a, the developmental path between different time points
is difficult to observe due to a large number of sprawled out clusters in the
embedding. On the other hand, the developmental paths are easy to follow
between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
factor from $1$ to $4$.

@@

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases where cell identities are not known in
advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
embedding using annotated data from Hochgerner \textit{et
al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
cells from the mouse brain. The authors provide annotations for each cell, and
we can verify their accuracy by visualizing well-established gene markers for the
major cell types. We then embed cells from Harris \textit{et
al.}~\cite{harris2018classes}, where annotations are provided only for neuronal
cells. We quickly identify other non-neuronal cell types, including
oligodendrocytes and astrocytes. We use marker genes to validate that new cells
were correctly mapped into the reference landscape.

@@

These two examples demonstrate how \opentsne\ can be used to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing already
existing cell atlases. This approach is extremely
general and is not limited to single-cell gene-expression data, but can be
applied to any tabular data sets from any domain of science.

@@

Versatility is another of  \opentsne s core design principles. Kobak \& Berens
recently introduced several recommendations and tricks to obtain better and
more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
of multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with larger data sets. \opentsne\ provides an
extremely flexible API, where these improvements can be incorporated in just a
few lines of code. Furthermore, \opentsne\ supports fully-custom affinity models,
enabling users to construct t-SNE embeddings on
non-tabular data, e.g. relational data, as the only affinity-model requirement
is some notion of similarity between data points. Finally, our comprehensive
callback system can be utilized to monitor and adapt different stages of the
optimization phase and has been used to construct visually appealing animations
of the t-SNE optimization process.

@@

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
\opentsne\ exhibit better scaling than their Barnes-Hut
counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
even though \textsf{FIt-SNE} and \opentsne\ both implement the same
approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
written in Python, which inevitably incurs some runtime overhead compared to
its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
than its Barnes-Hut competitors. Additionally, modern computer processors
generally contain multiple cores, allowing us to make use of multi-threading.
This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart when utilizing multiple cores. We speculate that this is could be because \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which is aggressively optimized on
Intel(R) Xeon(R) processors.

@@

This section is meant to familiarize the reader with the t-SNE algorithm and
its recent extensions. We motivate each extension and introduce standard notation.
This brief review is by no means comprehensive, but is
self-contained and introduces the mathematics upon which the \opentsne\ library
is based.
